---
title: "HW 1  - Parts 2 and 3: Instructions"
author: "Penelope Pooler Eisenbies"
date: "Due 9/7/2022"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=F}

# this line specifies options for default options for all R Chunks
knitr::opts_chunk$set(echo=T, 
                      highlight=T,
                      scipen=100)

# install helper package (pacman) if needed
if (!require("pacman")) install.packages("pacman", repos = "http://lib.stat.cmu.edu/R/CRAN/")

# install and load required packages
# pacman should be first package in parentheses and then list others
pacman::p_load(pacman, tidyverse, gridExtra, magrittr)

# verify packages (comment out in finished documents)
p_loaded()

```

## Purpose

This assignment will give you experience with:

- Creating an R Project Directory

- Creating a code_data_output folder within your R project to store files

- Creating, saving, and using an R Markdown file

- Reviewing some basic R syntax

---

## NOTES:

- Familiarity with R and R Studio infrastructure, including R Projects and R Markdown, is necessary for you to succeed in this course.

- Throughout this course you will be managing and manipulating data for different goals.

- You will not be tested directly on file management, but you can not manage data well until you can manage files well on your computer.

- File management details are not trivial. 

   - Good file management is necessary for good data management and will make all of your work more efficient. 

***

## Instructions

### HW 1 - Parts 2 and 3 - First Steps

0. Create a BUA455 folder on your computer, NOT in Downloads.

  - ALL work for this class will be saved to this folder 
  
  - My folder is named: **`00_BUA_455_All`**
  
<br>

1. Create and name your first R project for your first HW Assignment:

   i. File > New Project > New Directory > New Project 
   
   ii. In box, name this project: 
   
       - **`HW 1 <first name> <last name>`**
    
       - My project would be named 
        
           **`HW 1 Penelope Pooler`**
    
   iii. Click Create Project
   
<br>

2. Create a folder in HW 1 project labeled **`code_data_output`**.

   i. With Files tab open in R, click folder with green plus sign to add a folder to your project

   ii. Name the new folder **`code_data_output`** (all lower case with no spaces)


   iii. Click OK to create new folder

   iv. Verify that new folder is appears in Files window then close RStudio.

   v. Reopen the project by clicking on .Rproj file in the R project directory.

      - You do not have close and re-open RStudio every time.
   
      - This first time it is helpful to test that your project and **code_data_output`** have been correctly created.

<br>


3. Create a NEW R Markdown file with title HW 1 and you as the Author.
  

4. Save this file as 'HW1_FirstName_LastName' to your code_data_output folder.

    - Markdown files (.Rmd) should have no spaces in the file name.

<br>

5. Copy and paste the text shown below into your setup chunk.


6. Click the green triangle or type Ctrl/Cmd + Shift + Enter to run this setup chunk.
   
    - This was demonstrated in the Week 1 Lectures

```
# this line specifies options for default options for all R Chunks
knitr::opts_chunk$set(echo=T, 
                      highlight=T,
                      scipen=100)

# install helper package (pacman) if needed
if (!require("pacman")) install.packages("pacman", repos = "http://lib.stat.cmu.edu/R/CRAN/")

# install and load required packages
# pacman should be first package in parentheses and then list others
pacman::p_load(pacman, tidyverse, gridExtra, magrittr)

# verify packages (comment out in finished documents)
p_loaded()

```

### HW 1 - Part 2

**Notes:**

- HW 1 - Part 1 includes the questions about the syllabus, hardware requirements and R/RStudio versions.

- Recall that in Week 1, we worked with the first default R Chunk that is included when you open an R Markdown file.

- In your **`HW 1`** Markdown file that you have created, you should see a default chunk that appears as shown here:

```{r cars}
summary(cars)
```

#### Steps to Follow

1. In your **`HW 1`** Markdown file that you have created, you will need delete the first chunk and create a new one OR replace the R code in this chunk with the code below.

- **Two ways to create a new chunk:**

   - Click green C or type `Ctrl+Alt+i` (Mac users use Cmd)
   
<br>

- The new chunk title should be **`examining the cars data`**. 
  
   - Add the title to the chunk header after r within the bracket: 
   
      - e.g. **`{r examining the cars data}`**
<br> 
  
- The text of this chunk (without the header) is shown below.

   - Recall that text after **`#`** symbols are comments and are ignored by R

   - Copy and paste this R code into your new R chunk:

```
# cars is an internal R dataset
# this code saves a copy of the cars data in the Global Environment
my_cars <- cars

# examine the dataset mycars using glimpse
glimpse(my_cars)

# same command with piping:
# read |> as 'is sent to' or 'goes into'
my_cars |>
  glimpse()

# select rows 3, 4 and 5 only:
my_cars[3:5,]

# select column 1 only:
my_cars[,1]

# select observations 10, 11, and 12 within column 1:
my_cars[10:12, 1]

# select observations 20, 30, and 40, within column 2:
my_cars[c(20,30,40),2]

```

---

2. Answer the following Blackboard questions using the code provided above and adding some code of your own as specified in the questions below.

<br>

#### Blackboard (BB) Questions

##### **BB Question 1**  

The **`cars`** dataset is a dataset that is internal to the R software.  The code above saves a **copy** of this dataset with a new name.

* Where is this copy dataset saved?

<br>
  
##### **BB Question 2**  

The name of this copy of the cars dataset is `____`.

- Type dataset name **EXACTLY** as it appears in the R code.

<br>

##### **BB Question 3**  

This copy of the cars dataset has   

- `____` rows (observations)

- `____` columns (variables)

<br>

##### **BB Question 4**  

The R command **`my_cars[3:5,]`** prints out rows       
`____`, `____`, and `____` of the **`my_cars`** dataset.

<br>

##### **BB Question 5**  

The command **`my_cars[,1]`** prints out column  
`____` of the **`my_cars`** dataset.

<br>

##### **BB Question 6**    

Complete the command **`my_cars[________, ____]`** to select **rows 12 through 15** of **column 2** of the **`my_cars`** dataset.

**Notes:**

- Rows to be selected are specified before the comma.

- Columns to be selected are specified after the comma.

- **`:`** is used to specify consecutive columns or rows in R.

- **Add your code to the chunk you created with the code above.**

   - Replace `___` with correct numbers and symbols in your code before submitting it in R.

   - Test your code to verify that it is correct.
   
   - You can click on the dataset in the **Global Environment** to view the full dataset and verify your work.
   
   - Your grade on HW 1 is based on both your submitted R project and your Blackboard questions.
   
<br>
  
##### **BB Question 7**   

Complete the command **`my_cars[c(________), ____]`** to select rows 10, 15, and 20 of column 1 of the **`my_cars`** dataset.

**Notes:**

- See notes above for Question 6

- Non-consecutive elements within **`c()`** are separated by commas

---

### HW 1 - Part 3

**Notes:**

- In the Week 1 lectures, we also worked with the **`starwars`** dataset.

- The **starwars** dataset is internal to the **`dplyr`** package (included in **`tidyverse`**) that we loaded in our **`setup`** chunk above. 

- Code from those R Chunks is included below and more is added.

   - **R code below will not run unless the `setup` chunk is run first.**
   
- These chunks are provided as a reference to show how to basic R data manipulations and summaries

- R code with and without piping is included as a reference

- In future work, most R code will be shown with piping only

---

#### Steps to Follow

1. In your **`HW 1`** Markdown file that you have created, you will need delete the second chunk.

2. Create **FIVE** new empty R chunks

- **Two ways to create a new chunk:**

   - Click green C or type `Ctrl+Alt+i` (Mac users use Cmd)
   
<br>

- The new chunk titles should be:

   - **`starwars 1`**   **`starwars 2`**   **`starwars 3`**   **`starwars 4`**   **`starwars 5`**
  
   - Add the titles to the chunk header after r within the bracket: 
   
      - e.g. **`{r starwars 1}`**
<br> 
  
- The text of each chunk (without the header) is shown below.

   - Recall that text after **`#`** symbols are comments and are ignored by R

   - Copy and paste this R code into your new R chunk:
   

---



#### R Code for **`starwars 1`** Chunk

```
# save a copy of starwars data to your global environment
my_starwars <- starwars

# examine the data with glimpse
glimpse(my_starwars)

# same glimpse command with piping
my_starwars |> glimpse()

```



#### R Code for **`starwars 2`** Chunk

```
# examine the list of species with unique command
# to specify a variable WITHIN a dataset we use $, the accessor operator
unique(my_starwars$species)

# same unique command using piping and pull command
my_starwars |> pull(species) |> unique()

# examine the list of haircolors
# again, we use $ to specify hair_color within this dataset
unique(my_starwars$hair_color)

```



#### R Code for **`starwars 3`** Chunk

```
# use table to summarize starwars data by sex and species
table(my_starwars$sex, my_starwars$hair_color)

# save this table as an object
sw_gender_hrclr_smry <- table(my_starwars$sex, my_starwars$hair_color)

```



#### R Code for **`starwars 4`** Chunk

```
# summarize height of starwars characters
summary(my_starwars$height)

# calculate the mean height (also included in summary above)
# NA's must be excluded with na.rm=T
mean(my_starwars$height, na.rm=T)

# same mean command with piping and pull command
my_starwars |> pull(height) |> mean(na.rm=T)

# NA's must also be excluded for min(), max(), median, sd(), etc.
sd(my_starwars$height, na.rm=T)

# same sd command with piping and pull command
my_starwars |> pull(height) |> sd(na.rm=T)

```



#### R Code for **`starwars 5`** Chunk

```
# summarize sex
summary(my_starwars$sex)

# summarize sex using table
table(my_starwars$sex)

# summarize sex but use as.factor
summary(as.factor(my_starwars$sex))

# save this last summary as an object
sw_sex_smry <- summary(as.factor(my_starwars$sex))

```

---

3. Answer the following Blackboard Questions:

<br>

#### Blackboard (BB) Questions

##### **BB Question 8**  

The name of this copy of the starwars dataset is `____`.

- Type dataset name **EXACTLY** as it appears in the R code.

<br>

##### **BB Question 9**  

This copy of the starwars dataset has     

- `_____` rows (observations)

- `_____` columns (variables)

<br>

##### **BB Question 10**  

The output from the **`glimpse(my_starwars)`** command lists each variable and shows its type.     

- This dataset has `_____` character variables (labeled ***`<chr>`***).

- This dataset has _____ numeric variables 
  
  - Numeric variables include both decimal (***`<dbl>`***) and integer (***`<int>`***) variables.
  
<br>

- **NOTES:** 
  
   - **`glimpse()`** is a newer alternative to the **`str()`** command.    
  
   - **`glimpse()`**will only work if the **`tidyverse`** package suite is installed and loaded.

<br>

##### **BB Question 11**  

The command **`table(my_starwars$sex, my_starwars$hair_color)`** prints out a summary table of the **`my_starwars`** data showing how many characters of each sex classification have each hair color. 

Use this summary to fill in the blanks below.

- Other than **`none`** and **`unknown`**, 
there are `_____` hair color categories.

  - **Note:** You could also use **`unique(my_starwars$hair_color)`** to answer this first question but not the next two.

<br>
  
- There are `_____` female characters with brown hair in this dataset.

- There are `_____` male characters with brown hair in this dataset.

<br>

##### **BB Question 12**  

The R command **`summary(my_starwars$sex)`** does not provide useful information because **`sex`** is a character variable in this dataset. 
  
Fill in the blanks below with just text (no spaces) that complete two options that can summarize the data in this variable:


- **Notes:**

  - The first option is the **same command used in Question 11**

  - The second option coerces the variable **`sex`** to be a factor using the R command **`as.factor`**.

  - These commands are included in the provided R code.

<br>
  
- **`_____(my_starwars$sex)`**


- **`summary(_____(my_starwars$sex))`**
 
---

### HW 1 - Parts 2 and 3 - Final Steps

1. Save your HW1 R Markdown File

- You are **NOT** required to knit this file to create an HTML file.
   
- You are welcome to try and we (myself and/or TA) can assist you.

<br>

2. Answer all 12 Blackboard questions associated with this assignment.

3. Zip your entire Project Directory into a compressed File and submit it.

- **NOTE:** 

  - In HW 2 and subsequent HW assignments, you will be asked to create a README file for your R Project, but that is **NOT** required for this assignment. 

***

## Grading Criteria

### Pre-class Survey (20%)

These required questions help me gauge your prior knowledge of R and other information that is helpful.

### HW 1 - Part 1 (20%)

Completing this set of Blackboard Questions is required and verifies your knowledge of the syllabus and the course hardware and software requirements.

### HW 1 - Parts 2 and 3 (60%)

The Blackboard (BB) Questions (1-12) and the submitted R project together comprise 60% of HW 1. 

<br>

- **(12 pts.)** Each Blackboard question for HW1 - Parts 2 and 3 is worth 1 point.  

- **(2 pts.)** Completing **HW 1 - First Steps** as specified.  

- **(2 pts.)** Part 2: Full credit for  
  
  - Correctly replacing default **`cars`** chunk R code with code provided above.
  
  - This chunk must have correct title in chunk header:
  
    - **`examining the cars data`**
    
  - Adding to the provided R code:
  
    - BB Questions 6 and 7 require that you write new commands that should be saved to your R chunk (titled **`examining the cars data`**).
    
<br>
    
- **(4 pts.)** Part 3: Full credit for

  - Correctly creating **FIVE** new R chunks with correct headers and provided code in each chunk.
  
<br>

- **(2 pts.)** Completing the **HW 1 - Final Steps** and correctly submitting your zipped HW1 project directory.

---



  



